Regression Cube Analysis of Cohort Study Data

### 2015-03-24 draft.tex
added all content to related work; needs restructuring

\section{Epidemiological Background} \label{MedicalAndTechnicalBackground}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiology unites experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} it is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition are extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is a epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS} are used to analyzed the data using the classical sequential epidemiological workflow, which uses images largely only to communicate results, rather than providing insight.
%%
A alternative data driven approach is described as follows.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data if too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
%\com{http://link.springer.com/chapter/10.1007/978-3-642-39146-0_1#page-1}
\com{Turkey Metrics Paper}
\\\\
The work of Piringer et al. \cite{Piringer}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
%%
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Angelelli et al. \cite{Angelelli} is similar to ours as they visualize both image-derived an non image data using cube data structures with focus on comparison and knowledge extraction.
%%
Their proposed engine represents multiple heterogenous cohort study data sets as multiple normalized data cubes to eliminate data redundancy between them and perform runtime measure aggregation when measures of the different cubes have to be combined.
%%
This allows for saving memory space.
%%
Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{pearssons R} to depict relationships towards target features and employ list views and scatter plots to visualize and rank them.
%%
We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
We employ the idea of fast conducting large scale correlation analysis between many features to derive insight into the data.

\paragraph{Visual Analysis in Public Health.}
\com{Problem: Only paper of co-authors in this section, will look one-sided!}
\com{Uli Niemann, Tommy Hielscher, Myra Spiliopoulou, Henry Völzke, and Jens-Peter Kühn. Can we classify the participants of a longitudinal epidemiological study from their previous evolution?.}

\com{Niemann, U.; Völzke, H.; Kühn, J.-P. Spiliopoulou, M. (2014), Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis, Expert Systems with Applications 41 (11) , 5405-5415 .}

\com{Glaßer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

@@

\com{Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies}

@@

\com{Regression Cube: A Technique for Multidimensional Visual Exploration and Interactive Pattern Finding}
\com{The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems}

\com{[Hierarchical Brushing of High-Dimensional Data Sets Using Quality Metrics]}

\com{Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization}

\paragraph{Visual Analysis of Heterogenous Data.}
Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
%
Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
%
Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
%
We employ the idea of adding variables via drag and drop into a canvas area.

Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
They are useful to gain an overview over numerous variables and their distributions.
Histograms, bar charts, scatter plots and heat maps are used to visualize variable combinations with regard to their type.
The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many variables (127 in our application scenario).
We incorporate the idea of adaptive type-dependent visualizations.
Dai et al. \cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatter plots with integrated regression lines.
Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypotheses and on geographical data.
We employ the use of small multiples for incorporating heterogenous data types for comparability.
Chui et al. \cite{Chui2011} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
While the work shows how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
% *********************** Our own Work ***********************
% Prior Work
\paragraph{Prior Work.}
We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
%
When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%
Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
%
Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
% *********************** / Our own Work ***********************
\section{Regression Cube Analysis of Cohort Study Data}
\com{Add Data Preprocessing Section.}
The basic idea of our \emph{Regression Cube} is to provide a overview visualization of large cohort study data sets towards target variables.
%% TODO BLIND
Overview visualizations of feature relationships as presented by Klemm et al. \cite{Klemm2014VIS} are often focused on relationships between the visualized features.
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
In epidemiology, these relationships are also of interest, but rather towards their explanatory power towards the target feature.
These target features often indicates the presence of the investigated disease.
As described in Section \ref{sec:RegressionAnalysis}, regression analyses are the statistical tool of choice for analyzing these relationships.
A regression model is based on expert knowledge, there is no unified rule on how to apply them to a given set of features, so they have to be applied with care.
\subsection{Cube Description using Regression Formula Notation}
Expert knowledge is introduced into regression analyses using the regression formulas.
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to 

@@

The statistical processors yield the second structure.
%%
They rely on the statistical programming language \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}.
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is a \texttt{R} package and provides a API for accessing it via HTTP calls.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
The backend functions necessary for all cube calculations are provided via an \texttt{R} package.
%%
It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
%%
The server workload balances are managed by the front-end code.

>>>>

\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiology unites experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} it is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition are extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is a epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS} are used to analyzed the data using the classical sequential epidemiological workflow, which uses images largely only to communicate results, rather than providing insight.
%%
A alternative data driven approach is described as follows.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data if too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

%% Regression Cube: A Technique for Multidimensional Visual Exploration and Interactive Pattern Finding
We use the term Regression Cube to refer to a comparative 3D representation of multiple regression models.
%%
Chan et al. \cite{Chan} use the same term to describe an extension of the 2D-Scatterplot representation of a linear regression models (incorporating solely metric features) to a 3D-Cube.
%%
They group subjects using a set of interaction techniques as well as clustering algorithms to calculate sub-groups, which can then be compared using their cube representation.
%%
Similar to \cite{Piringer}, they focus on highlighting details of the included models, rather than comparing models consisting of different features.
%%
Insight is derived by subject grouping, which spawns new cube correlations and therefore allows drilling down to the data.
%%
In contrast to this approach, we focus on comparing models using quality-of-fit measures and do not focus on subdividing the data sets to gain insight.
%%
We model expert knowledge through the definition of regression formulas, which is not the focus of Chan et al.
%%
\com{Does this suffice as reason on why we use the same term "Regression Cube" for different things?}

@@

Angelelli et al. \cite{Angelelli} is similar to ours as they visualize both image-derived an non image data using cube data structures with focus on comparison and knowledge extraction.
%%
Their proposed engine represents multiple heterogenous cohort study data sets as multiple normalized data cubes to eliminate data redundancy between them and perform runtime measure aggregation when measures of the different cubes have to be combined.
%%
This allows for saving memory space.
%%
Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's r} to depict relationships towards target features and employ list views and scatter plots to visualize and rank them.
%%
We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
We employ the idea of fast conducting large scale correlation analysis between many features to derive insight into the data.

\paragraph{Visual Analysis in Public Health.}
\com{Uli Niemann, Tommy Hielscher, Myra Spiliopoulou, Henry Völzke, and Jens-Peter Kühn. Can we classify the participants of a longitudinal epidemiological study from their previous evolution?.}
%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al. \cite{Niemann2014} investigate risk factors of the hepatic steatosis (fatty liver disease) using decision trees with a interactive data mining tools.
%%
They extracted association rules, which serve as basis for this works proof-of-concept evaluation.
%%
We try to reproduce their results and investigate their findings further.
%%
\com{Glaßer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

@@

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al. \cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates and time plots to display and select data.
%%
Medical image data can be incorporated to investigate potentially interesting individuals.
%%
We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.

@@

%\com{The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems}
Ahmadi et al. \cite{Ahmadi} define the \emph{Sparse Regression Cube}, which partitions sparse high-dimensional into subspaces, which are then described using by their most reliable linear regression model.
%%
In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on presenting a algebraic for efficient regression-model calculation to find the best fit for a subspace.
%%
The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

%% Hierarchical Brushing of High-Dimensional Data Sets Using Quality Metrics
Albuquerque et al. \cite{Albuquerque} present a interactive exploration framework displaying quality metrics of high dimensional data sets with brushing facilities to create subsets.
%%
These subsets are then analyzed in detail, following a drilling-down approach.
%%
They incorporate scatter plot matrices (SPLOMs) of quality metrics describing the features.
%%
We incorporate the idea of abstracting features for comparison.
%%
Turkay et al. \cite{Turkay} follow a similar approach by using both descriptive metrics for features as well as the features themselves and incorporate them into linked plots.
%%
The approach was applied to a cognitive aging studies, where new hypotheses could be derived.

% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al. \cite{Bertini} present a overview over quality metrics describing high dimensional data.
%%
%According to their systematization we analyze correlations.
%%
Their research agenda acts as guideline for our design, comprising application using domain-experts, perceptual tuning towards human pattern recognition of important aspects and scalability between different data set sizes.

\paragraph{Visual Analysis of Heterogenous Data.}
%%
Generalized Pairs plots (\texttt{GPLOM´S}) extend the idea of scatter plot matrices by pairwise visualizing heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter.
%%
Dai et al. \cite{Dai2005} incorporate small multiples using a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with Pearsson's r values.
%%
A \emph{concept map} summarized features related to a specified disease.
%%
Time-dependent epidemiological data is visualized by Chui et al. \cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender towards influenza- and salmonellosis-associated hospitalizations.
%%
We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%
\paragraph{Prior Work.}
In recent publications, we focused on analysis on characterization of the healthy aging process of the lumbar spine.
%%
Therefore, we utilized semiautomatic detection algorithm do detect the lumbar spine shape \cite{Klemm2013VMV}.
%%
The reliable part of the presented models are represented by the shape of the lumbar spine canal, which was extracted and described using line representation.
%%

We defined a Interactive Visual Analysis workflow for image-centric cohort study data, by extending the feature selection step (see \ref{sec:Background}) with a iterative analysis loop incorporating group selection and visualization using expert-input as well as clustering methods \cite{Klemm2014VIS}.
%%
Information visualizations of cohort study features were augmented with extracted image data, yielding linked views with both image-and non-image information.
%%
The popularity of the mosaic plot depicting \emph{Cram\'{e}r's V} contingency values with our domain experts was the inspiration for this work.
%%
The representation presented in \cite{Klemm2014VIS} only highlights pairwise correlation between features without a target disease.

%% TODO: Check DLPD for correct page numbers
While the presented techniques were well received with epidemiological experts, the explanatory power towards back pain was limited, yielding a analysis based on image-derived features, such as extracting and analyzing curvature, torsion and angle of the lumbar spine \cite{Klemm2015}.
%%
These measures were analyzed using a novel decision tree quality plot towards non-image data.
%%
We concluded that the model quality is not enough to characterize back pain.
\\\\
The difference of our presented approach towards our previous and other herein presented work is twofold:
\begin{enumerate}
	\item We focus on large scale analysis of a vast number of linear and logistic regression-models by assessing their quality-of-fit using descriptive metrics.
	\item The analysis is conducted towards a target feature and incorporates expert knowledge via the regression model definition rather than subdividing the underlying data.
\end{enumerate}
\com{Main Difference to our and other related work}
\section{Regression Cube Analysis of Cohort Study Data}
\com{Add Data Preprocessing Section.}
%%
The basic idea of our \emph{Regression Cube} is to provide a overview visualization of large cohort study data sets towards target variables.
%% TODO BLIND
Overview visualizations of feature relationships as presented by Klemm et al. \cite{Klemm2014VIS} are often focused on relationships between the visualized features.
%%
Correlation metrics, such as the \emph{Pearson product-moment correlation coefficient} or \emph{Cram\'{e}r's V} contingency values are incorporated to achieve this goal.
%%
In epidemiology, these relationships are also of interest, but rather towards their explanatory power towards the target feature.
%%
These target features often indicates the presence of the investigated disease.
%%
As described in Section \ref{sec:RegressionAnalysis}, regression analyses are the statistical tool of choice for analyzing these relationships.
%%
A regression model is based on expert knowledge, there is no unified rule on how to apply them to a given set of features, so they have to be applied with care.
%%
\subsection{Cube Description using Regression Formula Notation}
%%
Expert knowledge is introduced into regression analyses using the regression formulas.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target.
%%
%In order to calculate all combinations of two to three independent features, we can apply  \emph{Regression Cube} aims to 
%%

@@

The statistical processors yield the second structure.
%%
They rely on the statistical programming language \texttt{R}\footnote{Open Source; \href{http://r-project.org}{\texttt{r-project.org}}}.
%%
It is widely adopted in the statistical analysis community, yielding a rich support of fast state-of-the-art statistics algorithms as well newly published methods.
%%
\texttt{OpenCPU} is a \texttt{R} package and provides a API for accessing it via HTTP calls \cite{Ooms}.
%%
This way, any computer, which runs \texttt{R} can be turned into a statistical processor for our project.
%%
The backend functions necessary for all cube calculations are provided via an \texttt{R} package.
%%
It uses multi-core optimization to use all machine CPUs to speed up the calculation process.
%%
The server workload balances are managed by the front-end code.
